CA Affordabulity

### 2020-09-20 paper.Rmd
update code

## Profile of prevalence of private water utilities in CA (c. 2017))

There are ~ 2,800 Community Water Systems (defined as water systems serving at least 15 residential units on a year-round basis, generally applying to all municipal and institutional systems with residential components, but not commercial, school, and recreational facilities). Of these, 97.6% of the population is  served by the 653 systems that serve more than 3,300 people, while only 2.4% of the population is served by 2,173 small systems.

@@

A  caveat here is that this subset of utilties for the most part represents phenomenon in larger California utilities. Cost structures and dynamics affecting water pricing and privatization in rural areas may differ substantially from those in urban areas. Below, we see that the vast majority of utilities with rates available for analysis have service populations greater than 1,000.

>>>>

## Profile of prevalence of private water utilities in CA (c. 2017))
```{r ear}
inv <- read_excel("Data/EAR/inventory.xlsx",
                  sheet = "SDWISWaterSystemInventory071620") %>% filter(WaterSystemType=="C")

inv$size_cat[which(inv$Pop<=500)] = "1. Very Small (<500)" 
inv$size_cat[which(inv$Pop>500)] = "2. Small (501-3300)" 
inv$size_cat[which(inv$Pop>3301)] = "3. Medium (3301-10,000)" 
inv$size_cat[which(inv$Pop>10001)] = "4. Large (10,001- 100,000)" 
inv$size_cat[which(inv$Pop>100000)] = "5. Very Large (>100,000)"

size_summary <- inv %>% 
  group_by(size_cat) %>%
  add_tally() %>%
  summarise(count=mean(n),pop=sum(Pop)/10^6) %>%
  ungroup() %>%
  mutate(perc_pop=100*pop/sum(pop))
kable(size_summary, digits=1)

# ear13 <- read_tsv("Data/EAR/earsurveyresults_2013ry.zip",col_types="ccdcccdddc")
# ear14 <- read_tsv("Data/EAR/earsurveyresults_2014ry.zip",col_types="ccdcccdddc")
# ear15 <- read_tsv("Data/EAR/earsurveyresults_2015ry.zip",col_types="ccdcccdddc")
# ear16 <- read_tsv("Data/EAR/earsurveyresults_2016ry.zip",col_types="ccdcccdddc")
# ear17 <- read_tsv("Data/EAR/earsurveyresults_2017ry.zip",col_types="ccdcccdddc")
# ear18 <- read_tsv("Data/EAR/earsurveyresults_2018ry.zip",col_types="ccdcccdddc")
# 
# ear <- bind_rows(ear13,ear14,ear15,ear16,ear17,ear18)
# rm(ear13,ear14,ear15,ear16,ear17,ear18)
# save(ear,file="Data/EAR/ear.rds")
load("Data/EAR/ear.rds")

own <- ear %>% 
    filter(QuestionName == "Population" | 
                QuestionName == "Water System Classification" |
                QuestionName == "Water System Ownership" |
                QuestionName == "IsWholesaler") %>% 
    mutate(QuestionID=paste0("q",QuestionID)) %>%
    select(PWSID, Year, QuestionName,  QuestionResults) %>% 
    distinct(PWSID, Year, QuestionName, .keep_all=TRUE) %>%
    pivot_wider(names_from = QuestionName,
                values_from = QuestionResults) %>%
    group_by(PWSID) %>%
    mutate(maxYear=max(Year)) %>%
    ungroup() %>%
    filter(Year == maxYear) %>%
    filter(is.na(IsWholesaler)) %>%
 #   mutate(Population = as.numeric(Population)) %>%
    filter(`Water System Classification`=="Community Water System")

own_inv <- left_join(inv,own,by="PWSID")

There are  2,917 Community Water Systems (defined as water systems serving at least 15 residential units on a year-round basis, generally applying to all municipal and institutional systems with residential components, but not commercial, school, and recreational facilities). Of these, 97.6% of the population is  served by the 658 systems that serve more than 3,300 people, while only 2.4% of the population is served by 2,259
small systems.

@@

A  caveat here is that this subset of utilities for the most part represents phenomenon in larger California utilities. Cost structures and dynamics affecting water pricing and privatization in rural areas may differ substantially from those in urban areas. Below, we see that the vast majority of utilities with rates available for analysis have service populations greater than 1,000.
